Visualization

### 2021-03-09 analysis/05_model_metrics.Rmd
Debugging problem models, add fixed effect for year and scale variables.

metrics <- pins::pin_get("rarefied-medians", board = "github") %>%
  select(-cYEAR) %>%
  pivot_longer(., cols = setdiff(everything(), one_of("rarefyID", "STUDY_ID", "SamplePool", "SampleN", "num_years", "duration", "startYear", "endYear", "YEAR", "cell", "rarefied", "type", "realm", "climate", "habitat", "biome_map", "taxa", "organisms", "cent_lat", "cent_long", "abundance_type", "BROAD_TYPE", "taxa_mod", "climate_mod")), names_to = "metric", values_to = "value") %>%
  rename_with(tolower, .cols = setdiff(everything(), one_of("rarefyID")))  %>%
  mutate(logvalue = log(value + abs(floor(min(na.omit(value)))))) %>% # create a column for log transformation that adds the absolute value integer closest to the most negative value to create all positive values
  filter(!is.na(value))
```
####Data exploration

@@

Fit GLM to overall metric trend
```{r}
# model for overall trend (random effect w/varying intercept)
## should be GLS according to Dornelas supplement, but glmer isn't happy
metric_mean_fit <- metrics %>%
  mutate(minyear = min(year), maxyear = max(year)) %>%
  group_nest(metric, minyear, maxyear) %>%
  mutate(model = map(data, ~lmer(logvalue ~ year + (1|study_id) + (1|rarefyID), 
                                 data = .x, control = lmerControl(optimizer ="Nelder_Mead"))))

metric_mean_coef <- metric_mean_fit %>%
  mutate(coef = map(model, tidy)) %>% 
  #mutate(confint = map(model, confint)) %>%
  select(-data, -model) %>% 
  unnest(cols = c(coef))

mean_tests <- metric_mean_fit %>%
  #filter(metric %in% c("FRic", "FEve", "FDiv", "FDis", "S")) %>%
   mutate(p_value = map(model, ~drop1(.x, test = "Chisq") %>% select(`Pr(Chi)`) %>% slice(2) %>% pull())) %>% 
  select(metric, p_value) %>%
  left_join(metric_mean_coef %>% 
              filter(term == "year") %>% 
              select(metric, estimate))
```

@@

indiv_ts_coef <- bind_rows(indiv_ts_fit, indiv_ts_sing_fit) %>%
  mutate(coef = map(model, tidy)) %>% 
  select(-data, -model) %>% 
  unnest(cols = c(coef))
```

```{r}
fit_mod <- function(data){
  print(unique(data$rarefyID)) %>%
  lmer(logvalue ~ (1 + year|rarefyID), 
                                 data = data, control = lmerControl(optimizer ="Nelder_Mead"))
}

>>>>

metrics <- pins::pin_get("rarefied-medians", board = "github") %>%
  select(-cYEAR) %>%
  pivot_longer(., cols = setdiff(everything(), one_of("rarefyID", "STUDY_ID", "SamplePool", "SampleN", "num_years", "duration", "startYear", "endYear", "YEAR", "cell", "rarefied", "type", "realm", "climate", "habitat", "biome_map", "taxa", "organisms", "cent_lat", "cent_long", "abundance_type", "BROAD_TYPE", "taxa_mod", "climate_mod")), names_to = "metric", values_to = "value") %>%
  rename_with(tolower, .cols = setdiff(everything(), one_of("rarefyID")))  %>%
  mutate(logvalue = log(value + abs(floor(min(na.omit(value))))),
         year_scaled = scale(year)) %>% # create a column for log transformation that adds the absolute value integer closest to the most negative value to create all positive values
  filter(!is.na(value)) 
```
####Data exploration

@@

Visualize distributions of response
```{r}
metrics %>%
  filter(metric %in% c("FRic", "FEve", "FDiv", "FDis", "S")) %>%
  ggplot() +
  geom_histogram(aes(x = logvalue),
                 bins = 60) +
  facet_wrap(~metric, scales = "free") +
  labs(title = "Distribution of Log values")

metrics %>%
  filter(metric %in% c("FRic", "FEve", "FDiv", "FDis", "S")) %>%
  ggplot() +
  geom_histogram(aes(x = value),
                 bins = 60) +
  facet_wrap(~metric, scales = "free") +
  labs(title = "Distribution of raw values")
```

Fit GLM to overall metric trend

Convergence problems!! See here:
https://joshua-nugent.github.io/allFit/
https://rstudio-pubs-static.s3.amazonaws.com/33653_57fc7b8e5d484c909b615d8633c01d51.html
https://biologyforfun.wordpress.com/2018/04/09/help-i-have-convergence-warnings/

Maybe need something other than a gaussian distribution?

Model checks:
https://debruine.github.io/posts/normality/
performance::check_model()

```{r}
# model for overall trend (random effect w/varying intercept)
metric_mean_fit <- metrics %>%
  filter(metric %in%  c("FRic", "FEve", "FDiv", "FDis", "S_int")) %>%
  group_nest(metric) %>%
  mutate(model = map(data, ~lmer(logvalue ~ year_scaled + (1|study_id/rarefyID), data = .x,
                  control = lmerControl(optCtrl = list(maxit = 1e9,
                                                    maxfun = 1e9,
                                                    xtol_abs = 1e-11,
                                                    ftol_abs = 1e-11,
                                                    maxeval = 1e9)))),
         null_model = map(data, ~lmer(logvalue ~ (1|study_id/rarefyID), data = .x,
                  control = lmerControl(optCtrl = list(maxit = 1e9,
                                                    maxfun = 1e9,
                                                    xtol_abs = 1e-11,
                                                    ftol_abs = 1e-11,
                                                    maxeval = 1e9)))),

           test = map2(model, null_model, anova),
           test_res = map(test, tidy),
           coef = map(model, tidy)
           ) %>% 
  unnest(cols = c(test_res)) %>%
  select(metric, model, coef, p.value) %>%
  filter(!is.na(p.value)) %>%
  unnest(cols = c(coef)) %>%
  filter(term == "year_scaled") %>%
  select(-c(effect, group, term, statistic))

metric_mean_fit_slope <- metrics %>%
  filter(metric %in%  c("FRic", "FEve", "FDiv", "S_int")) %>%
  group_nest(metric) %>%
  mutate(model = map(data, ~lmer(logvalue ~ year_scaled + (year_scaled|study_id/rarefyID), 
                                 data = .x)),
           null_model = map(data, ~lmer(logvalue ~ (year_scaled|study_id/rarefyID), 
                                 data = .x)),
           test = map2(model, null_model, anova),
           test_res = map(test, tidy),
           coef = map(model, tidy)
           ) %>% 
  unnest(cols = c(test_res)) %>%
  select(metric, model, coef, p.value) %>%
  filter(!is.na(p.value)) %>%
  unnest(cols = c(coef)) %>%
  filter(term == "year_scaled") %>%
  select(-c(effect, group, term, statistic))
```

@@

indiv_ts_coef <- bind_rows(indiv_ts_fit, indiv_ts_sing_fit) %>%
  mutate(coef = map(model, tidy)) %>% 
  select(-data, -model) %>% 
  unnest(cols = c(coef))
```
Fit a single model to play around with 
```{r}
eve <- metrics %>% 
  filter(metric == "FEve")

feve_fit <- glmer(value ~ year + (1|study_id/rarefyID), data = eve, family = Gamma,
                  control = glmerControl(optCtrl = list(maxit = 1e9,
                                                    maxfun = 1e9,
                                                    xtol_abs = 1e-11,
                                                    ftol_abs = 1e-11,
                                                    maxeval = 1e9)))

#find best optimizer
diff_optims <- allFit(feve_fit, maxfun = 1e5)

optimx_options <- c("L-BFGS-B", "nlminb", "nlm", "bobyqa", "nmkb", "hjkb")

for(i in 1:length(optimx_options)){
  model_flex <- glmer(value ~ year + (1|study_id/rarefyID), data = eve, family = Gamma,
                     control = glmerControl(optimizer = "optimx",
                                           optCtrl = list(method = optimx_options[i],
                                                                   maxit = 1e9)))
  print(model_flex@optinfo$conv$lme4$messages)
}

algoptions <- c("NLOPT_LN_PRAXIS", "NLOPT_GN_CRS2_LM",
"NLOPT_LN_COBYLA", "NLOPT_LN_NEWUOA",
"NLOPT_LN_NEWUOA_BOUND", "NLOPT_LN_NELDERMEAD",
"NLOPT_LN_SBPLX", "NLOPT_LN_BOBYQA")

for(i in 1:length(algoptions)){
  model_flex <- glmer(value ~ year + (1|study_id/rarefyID), data = eve, family = Gamma,
                     control = glmerControl(optimizer = "nloptwrap",
                                           optCtrl = list(algorithm = algoptions[i],
                                                          maxeval = 1e7,
                                                          xtol_abs = 1e-9,
                                                          ftol_abs = 1e-9)))
  if(is.null(model_flex@optinfo$conv$lme4$messages)){
    print(paste0("One of the nloptwrap options, ", algoptions[i],", worked!"))
    print(summary(model_flex))
    break
  }
}

fit_mod <- function(data){
  print(unique(data$rarefyID)) %>%
  lmer(logvalue ~ (1 + year|rarefyID), 
                                 data = data, control = lmerControl(optimizer ="Nelder_Mead"))
}

### 2021-03-09 analysis/visualization.Rmd
Debugging problem models, add fixed effect for year and scale variables.

```{r}
usa <- map_data("usa")

plt <- ggplot() +
  geom_polygon(data = usa, aes(x=long, y = lat, group = group), fill = NA, color = "black") +
  coord_fixed(1.3)

plt +
  geom_point(data = biotime_data %>% filter(taxa == "Fish"), aes(x = longitude, y = latitude), color = "black", size = .25)

>>>>

```{r}
world <- map_data("world")

plt <- ggplot() +
  geom_polygon(data = world, aes(x=long, y = lat, group = group), fill = NA, color = "black", size = .25) +
  coord_fixed(1.3) +
  theme_minimal()

plt +
  geom_point(data = metrics %>% distinct(cent_lat, cent_long), aes(x = cent_long, y = cent_lat), color = "#cc5968", size = .75)

ggsave("study_map.png")
